---
title: "Investigating Pertussis Resurgence"
author: "Jazz Zhang A16149005"
format: pdf
---

```{r}
library(ggplot2)

# Q1
cdc <- data.frame(
                             Year = c(1922L,1923L,1924L,1925L,1926L,1927L,
                                      1928L,1929L,1930L,1931L,1932L,1933L,
                                      1934L,1935L,1936L,1937L,1938L,1939L,
                                      1940L,1941L,1942L,1943L,1944L,1945L,
                                      1946L,1947L,1948L,1949L,1950L,1951L,
                                      1952L,1953L,1954L,1955L,1956L,1957L,
                                      1958L,1959L,1960L,1961L,1962L,1963L,
                                      1964L,1965L,1966L,1967L,1968L,1969L,
                                      1970L,1971L,1972L,1973L,1974L,1975L,
                                      1976L,1977L,1978L,1979L,1980L,1981L,
                                      1982L,1983L,1984L,1985L,1986L,1987L,
                                      1988L,1989L,1990L,1991L,1992L,1993L,
                                      1994L,1995L,1996L,1997L,1998L,1999L,
                                      2000L,2001L,2002L,2003L,2004L,2005L,
                                      2006L,2007L,2008L,2009L,2010L,2011L,
                                      2012L,2013L,2014L,2015L,2016L,2017L,
                                      2018L,2019L,2020L,2021L),
     No..Reported.Pertussis.Cases = c(107473,164191,165418,152003,202210,
                                      181411,161799,197371,166914,172559,
                                      215343,179135,265269,180518,147237,214652,
                                      227319,103188,183866,222202,191383,
                                      191890,109873,133792,109860,156517,
                                      74715,69479,120718,68687,45030,37129,
                                      60886,62786,31732,28295,32148,40005,
                                      14809,11468,17749,17135,13005,6799,
                                      7717,9718,4810,3285,4249,3036,3287,
                                      1759,2402,1738,1010,2177,2063,1623,
                                      1730,1248,1895,2463,2276,3589,4195,
                                      2823,3450,4157,4570,2719,4083,6586,
                                      4617,5137,7796,6564,7405,7298,7867,
                                      7580,9771,11647,25827,25616,15632,10454,
                                      13278,16858,27550,18719,48277,28639,
                                      32971,20762,17972,18975,15609,18617,
                                      6124,2116)
   )

plot <- ggplot(cdc) +
  aes(Year, No..Reported.Pertussis.Cases) +
  geom_point() +
  geom_line() +
  labs(title = "Pertussis cases by year", y = "Number of reported pertussis cases")
```

```{r}
plot +
  geom_vline(xintercept = c(1946, 1996, 2019), linetype = "dotted", color = c("red", "blue", "purple"), size = 0.7)
```

Q2. Number of reported pertussis cases reduced dramatically after introduction of wP vaccine

Q3. Number of reported pertussis cases increased after switching to aP vaccine. Protection provided by the aP vaccine wanes faster than the wP vaccine

```{r}
library(jsonlite)
subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE) 
head(subject, 3)
```

```{r}
# Q4
table(subject$infancy_vac)
```

```{r}
# Q5
table(subject$biological_sex)
```

```{r}
# Q6
table(subject$biological_sex, subject$ethnicity)
```

```{r}
library(lubridate)
today()
```

```{r}
subject$age <- time_length( today() - ymd(subject$year_of_birth),  "years")

round(mean(subject$age[subject$infancy_vac=="wP"]))
round(mean(subject$age[subject$infancy_vac=="aP"]))

t.test(subject$age[subject$infancy_vac=="wP"], subject$age[subject$infancy_vac=="aP"])$p.value

```

Q7. i) 36; ii) 26; iii) they are significantly different

```{r}
# Q8
subject$age_at_boost <- time_length( ymd(subject$date_of_boost) - ymd(subject$year_of_birth),  "years")
```

```{r}
ggplot(subject) +
  aes(time_length(age, "year"),
      fill=as.factor(infancy_vac)) +
  geom_histogram(show.legend=FALSE) +
  facet_wrap(vars(infancy_vac), nrow=2) +
  xlab("Age in years")

t.test(subject$age_at_boost[subject$infancy_vac=="wP"], subject$age_at_boost[subject$infancy_vac=="aP"])$p.value
```

Q9. They are significantly different

```{r}
specimen <- read_json("https://www.cmi-pb.org/api/specimen", simplifyVector = TRUE) 
titer <- read_json("https://www.cmi-pb.org/api/plasma_ab_titer", simplifyVector = TRUE) 
```

```{r}
library(dplyr)

# Q9
meta <- inner_join(specimen, subject)
dim(meta)
```

```{r}
# Q10
abdata <- inner_join(titer, meta)
dim(abdata)
```

```{r}
# Q11
table(abdata$isotype)
```

```{r}
# Q12
table(abdata$dataset)
```

```{r}
igg <- abdata %>% filter(isotype == "IgG")
head(igg)
```

```{r}
# Q13
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot() + 
    xlim(0,75) +
  facet_wrap(vars(visit), nrow=2)
```

Q14. Pertussis toxin, pertactin, fimbriae, and filamentous hemagglutinin show differences in the level of IgG antibody over time. These are surface molecules of pertussis which the patients are vaccinated against. Tetanus toxoid and diphtheria toxin are toxins of a different species of bacteria, while OVA albumin is a control peptide. Patients receiving pertussis vaccine should not generate antibodies against these molecules. 

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac) +
  geom_boxplot(show.legend = FALSE) + 
  facet_wrap(vars(visit), nrow=2) +
  xlim(0,75) +
  theme_bw()
```

```{r}
igg %>% filter(visit != 8) %>%
ggplot() +
  aes(MFI_normalised, antigen, col=infancy_vac) +
  geom_boxplot(show.legend = FALSE) + 
  xlim(0,75) +
  facet_wrap(vars(infancy_vac, visit), nrow=2)
```


```{r}
# Q15
filter(igg, antigen=="OVA") %>%
  ggplot() +
  aes(MFI_normalised, col=infancy_vac) +
  geom_boxplot(show.legend = TRUE) +
  facet_wrap(vars(visit)) +
  theme_bw() +
  labs(title = "OVA antigen level per visit")
```

```{r}
filter(igg, antigen=="FIM2/3") %>%
  ggplot() +
  aes(MFI_normalised, col=infancy_vac) +
  geom_boxplot(show.legend = TRUE) +
  facet_wrap(vars(visit)) +
  theme_bw() +
  labs(title = "FIM2/3 antigen level per visit")
```

Q16. IgG level against FIM2/3 increased over time but not against OVA. The trend is similar between aP and wP responses.

Q17. There is no clear difference between aP and wP responses

```{r}
abdata.21 <- abdata %>% filter(dataset == "2021_dataset")

abdata.21 %>% 
  filter(isotype == "IgG",  antigen == "PT") %>%
  ggplot() +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```

```{r}
abdata.20 <- abdata %>% filter(dataset == "2020_dataset")

abdata.20 %>% 
  filter(isotype == "IgG",  antigen == "PT") %>%
  ggplot() +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2020 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")+
  xlim(0, 125)
```

Q18. They don't look the same, the 2021 dataset show wP response show higher antibody titer on day 14 but not in 2020 dataset

```{r}
abdata.22 <- abdata %>% filter(dataset == "2022_dataset")

abdata.22 %>% 
  filter(isotype == "IgG",  antigen == "PT") %>%
  ggplot() +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2022 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```

```{r}
url <- "https://www.cmi-pb.org/api/v2/rnaseq?versioned_ensembl_gene_id=eq.ENSG00000211896.7"

rna <- read_json(url, simplifyVector = TRUE) 
ssrna <- inner_join(rna, meta)
```

```{r}
# Q19
ggplot(ssrna) +
  aes(visit, tpm, group=subject_id) +
  geom_point() +
  geom_line(alpha=0.2)
```

Q20. The expression level of this gene peaks at the 4th visit and drops after

Q21. It does not match the antibody production. Antibody level peaks at the 5th visit and is maintained at similar level until the 6th visit. Antibody needs more than to be translated into protein from mRNA, and since antibodies have relatively long half-lives, they can accumulate and are maintained in serum longer than mRNA. 

```{r}
ggplot(ssrna) +
  aes(tpm, col=infancy_vac) +
  geom_boxplot() +
  facet_wrap(vars(visit))
```

```{r}
ssrna %>%  
  filter(visit==4) %>% 
  ggplot() +
    aes(tpm, col=infancy_vac) + geom_density() + 
    geom_rug() 
```

