---
title: "Lab 9 (Unsupervised Learning)"
author: "Jazz Zhang A16149005"
format: pdf
---

```{r}
wisc.df <- read.csv("WisconsinCancer.csv", row.names = 1)
head(wisc.df)
dim(wisc.df)
```

Q1. 569 observations

```{r}
diagnosis <- as.vector(wisc.df$diagnosis)
wisc.data <- wisc.df[,-1]
```

```{r}
sum(diagnosis=="M")
```

Q2. 212 malignant diagnosis

```{r}
sum(grepl("_mean", colnames(wisc.data)))
```

Q3. 10 variables end with `"_mean"`

```{r}
colMeans(wisc.data)

apply(wisc.data,2,sd)
```

```{r}
wisc.pr <- prcomp(wisc.data[,-31], scale=TRUE)
summary(wisc.pr)
```

Q4. 44.27% of the original variance 

Q5. 3 PCs are required

Q6. 7 PCs are required

```{r}
biplot(wisc.pr)
```

Q7. Nothing stands out, can't interpret

```{r}
plot(wisc.pr$x[,1], wisc.pr$x[,2], col = as.factor(diagnosis), 
     xlab = "PC1", ylab = "PC2")
```

```{r}
plot(wisc.pr$x[, 1], wisc.pr$x[, 3], col = as.factor(diagnosis), 
     xlab = "PC1", ylab = "PC3")
```

Q8. Data with different diagnosis is clustered majorly by PC1

```{r}
df <- as.data.frame(wisc.pr$x)
df$diagnosis <- diagnosis

library(ggplot2)

# Make a scatter plot colored by diagnosis
ggplot(df) + 
  aes(PC1, PC2, col=diagnosis) + 
  geom_point()
```

```{r}
pr.var <- wisc.pr$sdev^2
head(pr.var)
```

```{r}
pve <- pr.var/sum(pr.var)

plot(pve, xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```

```{r}
barplot(pve, ylab = "Precent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )
```

```{r}
# install.packages("factoextra")

library(factoextra)
fviz_eig(wisc.pr, addlabels = TRUE)
```

```{r}
wisc.pr$rotation["concave.points_mean",1]
```

Q9. -0.26, it negatively contributes to the first PC

```{r}
data.scaled <- scale(wisc.data)
data.dist <- dist(data.scaled)
wisc.hclust <- hclust(data.dist, method="complete")
```

```{r}
plot(wisc.hclust)
abline(h=19, col="red", lty=2)
```

Q10. height = 19

```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, h=19)
table(wisc.hclust.clusters, diagnosis)
```

```{r}
single <- hclust(data.dist, method="single")
plot(single)
```

```{r}
average <- hclust(data.dist, method="average")
plot(average)
```

```{r}
ward <- hclust(data.dist, method="ward.D2")
plot(ward)
```

Q12. `"ward.D2"` because variance is minimized within clusters using this 
method

```{r}
wisc.pr.hclust <- hclust(dist(wisc.pr$x[,1:7]), method="ward.D2")
grps <- cutree(wisc.pr.hclust, k=2)
table(grps)

table(grps, diagnosis)

plot(wisc.pr$x[,1:2], col=grps)
```

```{r}
plot(wisc.pr$x[,1:2], col=factor(diagnosis))
```

```{r}
g <- as.factor(grps)
levels(g)

g <- relevel(g,2)
levels(g)
```

```{r}
plot(wisc.pr$x[,1:2], col=g)
```

```{r}
# install.packages("rgl")

library(rgl)
plot3d(wisc.pr$x[,1:3], xlab="PC 1", ylab="PC 2", zlab="PC 3", cex=1.5, 
       size=1, type="s", col=grps)
```

```{r}
wisc.pr.hclust <- hclust(dist(wisc.pr$x[, 1:7]), method="ward.D2")
wisc.pr.hclust.clusters <- cutree(wisc.pr.hclust, k=2)

table(wisc.pr.hclust.clusters, diagnosis)
```

Q13. Clusters are majorly separated by diagnosis

```{r}
table(cutree(wisc.hclust, k=2), diagnosis)
```

```{r}
table(cutree(single, k=2), diagnosis)
```

```{r}
table(cutree(average, k=2), diagnosis)
```

```{r}
table(cutree(ward, k=2), diagnosis)
```

Q14. Diagnosis cannot be separated well with 2 clusters using the other methods
(`"single"`, `"average"`, and `"complete"`)

```{r}
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```

```{r}
plot(wisc.pr$x[,1:2], col=g)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```

Q16. Patient 2