---
title: "Halloween Mini-Project"
author: "Jazz Zhang (A16149005)"
format: pdf
---

```{r}
candy_file <- read.csv("candy-data.csv")
candy = read.csv("candy-data.csv", row.names=1)

dim(candy)
```

Q1. 85 candy types

```{r}
sum(candy$fruity)
```

Q2. 38 fruity candy types

```{r}
candy["Dum Dums",]$winpercent
```

Q3. Dum Dums, 39.46%

```{r}
candy["Kit Kat",]$winpercent
```

Q4. 76.77%

```{r}
candy["Tootsie Roll Snack Bars",]$winpercent
```

Q5. 49.65%

```{r}
# install.packages("skimr")
library("skimr")
skim(candy)
```

Q6. "winpercent" column

Q7. 0 and 1 represent the candy is either chocolate or not, repectively

```{r}
library(ggplot2)

# Q8.
ggplot(candy, aes(winpercent))+
  geom_histogram()
```

Q9. The distribution isn't symmetrical

Q10. The ccenter is below 50

```{r}
t.test(candy$winpercent[as.logical(candy$chocolate)], y=candy$winpercent[as.logical(candy$fruity)])
```

Q11. Winpercent for chocolate is higher on average

Q12. The difference is statistically significant

```{r}
library(dplyr)

candy %>% arrange(winpercent) %>% head(5)
```

Q13. Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, and Jawbusters

```{r}
candy %>% arrange(desc(winpercent)) %>% head(5)
```

Q14. Reese's Peanut Butter cup, Reese's Miniatures, Twix, Kit Kat, and Snickers

```{r}
# Q15.
ggplot(candy) + 
  aes(x=winpercent, y=rownames(candy)) +
  geom_bar(stat="identity")
```

```{r}
# Q16.
ggplot(candy) + 
  aes(x=winpercent, y=reorder(rownames(candy),winpercent)) +
  geom_bar(stat="identity")
```

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```

```{r}
ggplot(candy) + 
  aes(x=winpercent, y=reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

Q17. Sixlets

Q18. Starburst

```{r}
library(ggrepel)

ggplot(candy) +
  aes(x=winpercent, y=pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 50)
```

Q19. Reese's Miniatures

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

Q20. Nik L Nip, Nestle Smarties, Ring pop, Hershey's Krackel, and Hershey's 
Milk Chocolate; Nik L Nip is the least popular

```{r}
# Q21.
ggplot(candy) + 
  aes(x=pricepercent, y=reorder(rownames(candy),pricepercent)) +
  geom_col() 
```

```{r}
# Q21. cont
ggplot(candy) + 
  aes(x=pricepercent, y=reorder(rownames(candy),pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
  geom_point()
```

```{r}
# install.packages("corrplot")

library(corrplot)

cij <- cor(candy)
corrplot(cij)
```

Q22. Chocolate and fruity

Q23. Chocolate and winpercent

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

```{r}
plot(pca$x[,1:2])
```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

```{r}
my_data <- cbind(candy, pca$x[,1:3])

p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 50)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
#| eval: false
library(plotly)

ggplotly(p)
```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

Q24. Fruity, pluribus, and hard. 